---
title: Genome-wide analyses of cell-free DNA for therapeutic monitoring of patients with pancreatic cancer
author: Carolyn Hruban
output:
  md_document:
    variant: markdown_github
---

# README

The data used in this analysis originates from the [CheckPAC](https://www.clinicaltrials.gov/study/NCT02866383) and [PACTO](https://clinicaltrials.gov/study/NCT02767557) trials. 

The CRAM files associated with these datasets have been deposited in the European Genome-phenome Archive (EGA):

- **Study ID** EGAS50000000923
- **Dataset ID** EGAD50000001353

## Overview

This repository was used to generate the [figures](https://cancer-genomics.github.io/hruban_wflow/index.html) for the associated manuscript.

## Reproducing the Figures

Clone the Repository:
```{sh, eval = FALSE}
git clone https://github.com/cancer-genomics/hruban_wflow
```

Navigate to the Repository Directory:
```{sh, eval = FALSE}
cd hruban_wflow
```

Render the README File (ensure that eval is set to TRUE in the code chunks):
```{sh, eval = FALSE}
Rscript -e 'rmarkdown::render("README.Rmd")'
```
  
## Required Packages

The following R packages are required for this analysis. Setting eval = TRUE during the rendering process will ensure they are installed:

```{r install_packages, message = FALSE, warning = FALSE, eval = FALSE}
options(repos = structure(c(CRAN = "getCRANmirrors()$URL[1]")))
BiocManager::install("ComplexHeatmap")
BiocManager::install("VplotR")
install.packages("RColorBrewer")
install.packages("caret")
install.packages("circlize")
install.packages("cowplot")
install.packages("data.table")
install.packages("forcats")
install.packages("fs")
install.packages("glmnet")
install.packages("ggpubr")
install.packages("ggstance")
install.packages("gridExtra")
install.packages("gridGraphics")
install.packages("here")
install.packages("lubridate")
install.packages("matrixStats")
install.packages("openxlsx")
install.packages("plotly")
install.packages("readxl")
install.packages("recipes")
install.packages("rstatix")
install.packages("scales")
install.packages("survival")
install.packages("survminer")
install.packages("tidyverse")
install.packages("viridis")
```

## Pipelines Used

The pipelines used in this publication are either included in this repository or have been published previously:

- **WGMAF** - Located in `code/wgmaf_code` 
- **DELFI pipeline** - [GitHub Repository](https://github.com/cancer-genomics/delfi3)
- **ARTEMIS-DELFI model** - [Annapragada, AV. et al (2024)](https://www.science.org/doi/10.1126/scitranslmed.adj9283)
- **DELFI TF model** - [van’t Erve, I. et al (2024)](https://doi.org/10.1038/s41467-024-53017-7)

## Directory Structure
This repository follows the [workflowr](https://github.com/workflowr/workflowr) project structure.

`analysis/` - Contains the scripts needed to reproduce the manuscript figures. These can be rebuilt using the following command:

```{r rebuild_figs, eval = FALSE}
library(workflowr)
wflow_build("analysis/*.Rmd")
```

`code/` - Contains scripts required to generate intermediate objects that are later loaded by the `analysis/` scripts. Some of these may need to be run on a high-performance computing cluster.

`data/` - Contains intermediate data objects used in the analysis.

`extdata/` - Not included due to size restrictions. This directory contains two files:

`hic_compartments_100kb_ebv_2014.txt`: Downloaded from [GitHub repository](https://github.com/Jfortin1/HiC_AB_Compartments)

`paad_betas_100kb.csv`: PDAC A/B compartments were generated from TCGA methylation data according to [Fortin, JP. et al (2015)](https://doi.org/10.1186/s13059-015-0741-y) 
