---
title: "Combine TCGA bins with CheckPAC sample coverages / ratios"
author: "Carlie"
date: "`r Sys.Date()`"
output: html_document
---

Load packages
```{r setup, include=FALSE}
library(tidyverse)
library(data.table)
library(here)

source(here("code/functions.R"))
sel_chr <- readLines(here("data/sel_chr.txt"))
indir <- here("output/process_tcga_beta.Rmd")
outdir <- here("output/process_ab.Rmd")
fs::dir_create(outdir)
```

Load TCGA bins, LCL and CheckPAC data
```{r load_data}
# Load processed TCGA bins
ref <- readRDS(here(indir, "paad_bins_100kb_sel_chr.rds"))
ref[, arm := NULL]  # Don't need it…

# Load LCL data
lymph <- fread(here("extdata/hic_compartments_100kb_ebv_2014.txt"))
lymph <- lymph[chr == sel_chr]

# Load CheckPAC data (WGMAF) and plasma ratios, coverage
ic <- fread(here("data/checkpac_meta_ab.csv"))
ic[, `:=`(tumor = delfi_tf, wbc = 1 - delfi_tf)]
ic <- ic[, .(id, wbc, tumor)]

cov_panc <- fread(here("data/plasma_panc.csv"))
cov_heal <- fread(here("data/plasma_healthy.csv"))

cov_panc_l <- melt(cov_panc, id.vars = 1, measure.vars = NULL)
cov_heal_l <- melt(cov_heal, id.vars = 1, measure.vars = NULL)

# Loading the genomic coordinates from a binned sample
#bins <- fread(here("data/CGPLPA223P_100kb.csv"))[, .(chr, start, end, bin)]
bins <- fread(here("data/bins_100kb.csv"))[, .(chr, start, end)]
# I did not notice that lymphoblast and TCGA bins had different coordinates
# fixing it here :(
bins[, `:=`(bin = 1:.N, start = start - 1, end = end - 1)]
bins <- bins[chr != "chrX"]
```

Extract pancreatic component using CheckPAC and healthy plasma samples
```{r compute_pancan}
# Compute median values not using na.rm = TRUE, NAs are set to 0
# For CheckPAC samples
median_panc_coverage <- cov_panc_l[grep("cov_", variable),
                                   .(value = median(value)),
                                   by = variable]
median_panc_coverage[, value := nafill(value, fill = 0)]

median_panc_ratio <- cov_panc_l[grep("ratio_", variable),
                                .(value = median(value)),
                                by = variable]
median_panc_ratio[, value := nafill(value, fill = 0)]

# For Healthy samples
median_heal_coverage <- cov_heal_l[grep("cov_", variable),
                                   .(value = median(value)),
                                   by = variable]
median_heal_coverage[, value := nafill(value, fill = 0)]

median_heal_ratio <- cov_heal_l[grep("ratio_", variable),
                                .(value = median(value)),
                                by = variable]
median_panc_ratio[, value := nafill(value, fill = 0)]

# "Deconvolve" cancer signal
deconv_cov_panc_l <- ic[cov_panc_l[median_heal_coverage, on = .(variable)], on = .(id)]
median_deconv_coverage <- deconv_cov_panc_l[, .(value = median((value - wbc * i.value) / tumor)), by = variable]

deconv_ratio_panc_l <- ic[cov_panc_l[median_heal_ratio, on = .(variable)], on = .(id)]
median_deconv_ratio <- deconv_ratio_panc_l[, .(value = median((value - wbc * i.value) / tumor)), by = variable]
```

Assign bins to CheckPAC, healthy and extracted pancreatic components
```{r bin_assign}
panc_list <- list(Ratios = cbind(bins, median_panc_ratio)[, eigen := scale(value, scale = TRUE)[, 1]],
                  Coverage = cbind(bins, median_panc_coverage)[, eigen := scale(value, scale = TRUE)[, 1]])
panc_tbl <- rbindlist(panc_list, idcol = "feature")

healthy_list <- list(Ratios = cbind(bins, median_heal_ratio)[, eigen := scale(value, scale = TRUE)[, 1]],
                     Coverage = cbind(bins, median_heal_coverage)[, eigen := scale(value, scale = TRUE)[, 1]])
healthy_tbl <- rbindlist(healthy_list, idcol = "feature")

deconv_list <- list(Ratios = cbind(bins, median_deconv_ratio)[, eigen := scale(value, scale = TRUE)[, 1]],
                    Coverage = cbind(bins, median_deconv_coverage)[, eigen := scale(value, scale = TRUE)[, 1]])
deconv_tbl <- rbindlist(deconv_list, idcol = "feature")
```

Assign bins to `ref` and `lymph`
```{r bin_assign_ref}
ref_bin <- bins[ref, on = .(chr, start, end)]
lymph_bin <- bins[lymph, on = .(chr, start, end)]
```

Smoothen the data and update signs if "neccessary"
```{r smooth_sign}
#ref_bin[, eigen := .meanSmoother(eigen)]   # If > 1 chr, use by = chr 
deconv_tbl[, eigen := -1 * .meanSmoother(eigen), by = .(feature, chr)]
healthy_tbl[, eigen := -1 * .meanSmoother(eigen), by = .(feature, chr)]
panc_tbl[, eigen := -1 * .meanSmoother(eigen), by = .(feature, chr)]
#lymph_bin[, eigen := .meanSmoother(eigen)] # If > 1 chr, use by = chr 
```

Assign open-closed domains to above components
```{r label_AB}
panc_tbl[, domain := ifelse(eigen < 0, "open", "closed")]
panc_tbl[, domain := factor(domain)]

healthy_tbl[, domain := ifelse(eigen < 0, "open", "closed")]
healthy_tbl[, domain := factor(domain)]

deconv_tbl[, domain := ifelse(eigen < 0, "open", "closed")]
deconv_tbl[, domain := factor(domain)]
```

Mark if domains agree with `ref` and `lymph`
```{r domain_agreement}
ref_bin <- ref_bin[lymph, on = .(chr, start, end)][, lymph_agree := ifelse(domain == i.domain, "yes", "no")][, lihc_agree := lymph_agree]
ref_bin[, .(chr, start, end, bin, nprobes, eigen, domain, lihc_agree, lymph_agree)]
ref_bin <- na.omit(ref_bin)

panc2 <- panc_tbl[ref, on = .(chr, start, end)][, lihc_agree := ifelse(domain == i.domain, "yes", "no")]
panc2 <- panc2[, .(feature, chr, start, end, bin, variable, value, eigen, domain, lihc_agree)]
panc3 <- panc2[lymph, on = .(chr, start, end)][, lymph_agree := ifelse(domain == i.domain, "yes", "no")]
panc4 <- panc3[, .(feature, chr, start, end, bin, variable, value, eigen, domain,
                   lihc_agree, lymph_agree)]
panc <- na.omit(panc4)

healthy2 <- healthy_tbl[ref, on = .(chr, start, end)][, lihc_agree := ifelse(domain == i.domain, "yes", "no")]
healthy2 <- healthy2[, .(feature, chr, start, end, bin, variable, value, eigen, domain, lihc_agree)]
healthy3 <- healthy2[lymph, on = .(chr, start, end)][, lymph_agree := ifelse(domain == i.domain, "yes", "no")]
healthy4 <- healthy3[, .(feature, chr, start, end, bin, variable, value, eigen, domain,
                         lihc_agree, lymph_agree)]
healthy <- na.omit(healthy4)

deconv2 <- deconv_tbl[ref, on = .(chr, start, end)][, lihc_agree := ifelse(domain == i.domain, "yes", "no")]
deconv2 <- deconv2[, .(feature, chr, start, end, bin, variable, value, eigen, domain, lihc_agree)]
deconv3 <- deconv2[lymph, on = .(chr, start, end)][, lymph_agree := ifelse(domain == i.domain, "yes", "no")]
deconv4 <- deconv3[, .(feature, chr, start, end, bin, variable, value, eigen, domain,
                       lihc_agree, lymph_agree)]
deconv <- na.omit(deconv4)
```

Combine data and mark domains that agree
```{r combine}
# Combine all data
data <- list(`AB Compartments PAAD` = ref_bin,
             `Extracted Pancreatic component` = deconv_tbl,
             `Median Healthy Plasma` = healthy_tbl,
             `Median Pancreatic Plasma` = panc_tbl,
             `EBV transformed lymphoblast HiC` = lymph_bin)
data_tbl <- rbindlist(data, idcol = "source", fill = TRUE)

# Label source2 = source + feature
data_tbl[, feature := ifelse(is.na(feature), "", feature)]
data_tbl[, source2 := str_trim(paste(source, feature))]

# Select needed columns and drop NAs
data_tbl <- data_tbl[, .(source2, source, feature, chr, start, end, bin,
                         nprobes, eigen, domain, lihc_agree, lymph_agree)]
data_tbl <- na.omit(data_tbl, cols = "eigen")

# Keep same bins across data sets (N = 8)
common_bins <- data_tbl[, .(nbin = .N), by = .(chr, start, end)]
common_bins <- common_bins[, .(keep = nbin == 8), by = .(chr, start, end)]
data_tbl <- data_tbl[common_bins[keep == TRUE], on = .(chr, start, end)]

# Fill the missing self-agreement on `ref` and `lymph` with "yes"
data_tbl[, `:=`(lihc_agree = ifelse(is.na(lihc_agree), "yes", lihc_agree),
                lymph_agree = ifelse(is.na(lymph_agree), "yes", lymph_agree))]

saveRDS(data_tbl, here(outdir, "combined_bins.rds"))
```
