---
title: Change in DELFI-TF scores between baseline and 8 week time point scores are predictive of progression free and overall survival in CheckPAC trial
---

Multivariate cox proportional hazard analyses were generated for DELFI-TF and fit to overall survival adjusting for clinical subgroups. Each of the indicated subgroups that have been shown to be significant on univariate analyses in previous studies(7) have been included in the multivariate analysis.  Hazard ratios and P values are indicated on the right for individuals in the CheckPAC study


```{r packages, message = FALSE}
library(here)
library(readxl)
library(survival)
source(here("code/hr_plot_funcs.R"))
```


```{r load_clin_data}
baseline_c <- read_excel("data/CheckPAC_clinical_annotated011622.xlsx") %>%
  select(subject_id = StudySubjectID,
         age,
         weightloss5_yn = weighttloss5_yn,
         ecog_ps = PS_cycle1,
         bor = `Best overall confirmed response (RECIST)`,
         ca199 = LAB_CA199_E1_C4,
         crp = LAB_CRP_E1_C4,
         bilirubin = LAB_Bilirubin_E1_C4,
         albumin = LAB_Albumin_E1_C4) %>%
  mutate(mgps = case_when(crp <= 10 ~ 0,
                          crp > 10 & albumin >= 35 ~ 1,
                          crp > 10 & albumin < 35 ~ 2))
long_c <- read_excel("data/clinical_073123.xlsx") %>%
  select(subject_id = StudySubjectID,
         arm = Arm,
         nivo_start_date = Nivo_StartDate,
         pfs_date = PFS_date,
         surv_date = Survival_date,
         surv_status = Survival_status) %>%
  mutate(surv_status = case_when(surv_status == 2 ~ "Dead",
                                 surv_status == 1 ~ "Alive"),
         arm = factor(arm))
clin_c <- inner_join(baseline_c, long_c, by = "subject_id")

# Fix the date formats
for (i in colnames(clin_c)) {
  if (str_detect(i, "[Dd]ate")) {
    if (!any(class(clin_c[[i]]) == "POSIXt")) {
      clin_c[[i]] <- as.Date(as.numeric(clin_c[[i]]), origin = "1899-12-30")
    } else {
      clin_c[[i]] <- as.Date(clin_c[[i]])
    }
  }
}

# Compute overall and progression-free survival time
surv_c <- clin_c %>%
  mutate(pfs = as.numeric(difftime(pfs_date, nivo_start_date)),
         os = as.numeric(difftime(surv_date, nivo_start_date))) %>%
  mutate(death = ifelse(surv_status == "Dead", 1, 0))
```

```{r load_delfi_tf}
# Load the DELFI TF scores
delfi_tf <- read_delim("data/df_results_CAIRO5_2_6_ checkpac_ pacto.tsv") %>%
  setNames(c("indx", "id", "score")) %>%
  select(-indx)

# Select baseline and two-week visit timepoints
blood_draw_c <- read_excel("data/checkpac_plasma_manifest.xlsx") %>%
  select(cgid = `Sample ID*`, patient_id = Patient, date_blood_draw = Date) %>%
  mutate(date_blood_draw = as.Date(date_blood_draw))
# See "/dcs05/scharpf/data/skoul/Projects/checkpac/patient_selection" for more
# information
blood_draw_c[blood_draw_c$cgid == "CGPLPA223P16",
             "date_blood_draw"] <- as.Date("2019-12-20")
blood_draw_c[blood_draw_c$cgid == "CGPLPA794P4",
             "date_blood_draw"] <- as.Date("2020-02-21")
blood_draw_bl_or_p2_c <- blood_draw_c %>%
  mutate(visit = str_extract(cgid, "(?<=CG[A-Z]{1,5}[0-9]{1,5}P).*"),
         visit = sapply(visit, function(x) str_split(x, "_")[[1]][1])) %>%
  filter(visit == "" | visit == "2")
blood_draw_bl_and_p2_c <- blood_draw_bl_or_p2_c %>%
  filter(patient_id %in% pull(filter(count(group_by(blood_draw_bl_or_p2_c,
                                                    patient_id)),
                                     n > 1),
                              patient_id))

# Join the scores with timepoint
delfi_tf_blp2_c <- inner_join(delfi_tf, blood_draw_bl_and_p2_c,
                              by = join_by(id == cgid)) %>%
  mutate(visit = ifelse(visit == "", "baseline", "ontreatment"))
```


```{r combine_clin_delfi_tf}
exclude_c <- c("CGPLPA225", "CGPLPA783", "CGPLPA795")
id_map_c <- read_excel("data/ID_comparison.xlsx") %>%
  filter(!unlist(lapply(CGID, function(x) any(str_detect(x, exclude_c))))) %>%
  select(patient_id = CGID, subject_id = StudySubjectID) %>%
  mutate(patient_id = substring(patient_id, 1, 9))

surv_delfi_tf_c <- surv_c %>%
  inner_join(id_map_c) %>%
  inner_join(delfi_tf_blp2_c) %>%
  select(subject_id, patient_id, score,
         arm, ca199, bilirubin, age, mgps, ecog_ps, weightloss5_yn, visit,
         pfs, os, death) %>%
  pivot_wider(names_from = visit, values_from = score) %>%
  mutate(bilirubin_bin = bilirubin > 25,
         age_bin = age <= 65,
         mgps_bin = factor(mgps, levels = c(0, 1, 2)),
         ecog_ps_bin = ecog_ps == 1,
         ca199_bin = ca199 > 37,
         weightloss5_yn_bin = weightloss5_yn == 1)

# Data for fitting the models
surv_data <- surv_delfi_tf_c %>%
  mutate(pch = (ontreatment - baseline) / baseline,
         pch_gt0 = ifelse(pch > 0, "NR", "R"),
         baseline_std = scale(baseline, center = TRUE, scale = TRUE),
         ontreatment_std = scale(ontreatment, center = TRUE, scale = TRUE))
```


```{r ext-fig21b, fig.width = 12, fig.height = 5, dev = c("png", "pdf")}
covar_labels <- c("Age, years",
                  "> 65",
                  "\u2264 65",
                  "ECOG PS",
                  "0",
                  "1",
                  "Weight loss, %",
                  "< 5",
                  "\u2265 5",
                  "MGPS",
                  "0",
                  "1",
                  "2",
                  "Treatment",
                  "RT+Nivo",
                  "RT+Nivo/Ipi")

# Inc-dec model
var_labels <- c("DELFI TF",
                "Non-responder",
                "Responder")

# Fit overall survival to inc-dec Delfi TF values
mf <- coxph(Surv(os, death) ~ age_bin + ecog_ps_bin + weightloss5_yn_bin +
              mgps_bin + arm + pch_gt0, surv_data)
cox_os <- get_hr_tbl(surv_data, mf)
print(plot_hr(cox_os, var_labels = c(covar_labels, var_labels)))
```


```{r ext-fig21b-pfs, fig.width = 12, fig.height = 5, dev = c("png", "pdf"), eval = FALSE}
# Fit progression-free survival to inc-dec Delfi TF values
mf <- coxph(Surv(pfs, death) ~ age_bin + ecog_ps_bin + weightloss5_yn_bin +
              mgps_bin + arm + pch_gt0, surv_data)
cox_pfs <- get_hr_tbl(surv_data, mf)
print(plot_hr(cox_pfs, var_labels = c(covar_labels, var_labels)))
```
