---
title: Timepoints selected for CA19-9 and CT for patients in CheckPAC trial 
---


Blue lines indicate the date of CT. Red-X indicates patients dropped due to missing sample. The highlighted blue region is the 0.05-0.95 Quantiles of timepoint selection. Fourth follow up CA19-9 timepoints were selected as follow-up, which are outlined in black. In the legend, timepoint (baseline, 1-5 or NA) indicates follow up CA19-9 sample number. 


```{r load_packages, message = FALSE}
library(data.table)
library(here)
library(lubridate)
library(readxl)
library(tidyverse)
source(here("code/utility_funcs.R"))
```

**NOTE**: it's better to move this code to `code` and refer to its output in analysis
and drop the copy from data
```{r load_data}
# Load the master ID map file - from Carlie
id_map <- read_excel(here("data/ID_comparison.xlsx")) %>%
  select(cgid = CGID, subject_id = StudySubjectID) %>%
  mutate(patient_id = gsub("P[0-9]{0,2}$", "", cgid)) %>%
  select(-cgid)

# Load the blood draw dates
mfst <- read.csv(here("data/checkpac_ca199.csv")) %>%
  mutate(visit_cycle = ifelse(lab_suffix == "E1_C4",
                              "baseline",
                              as.character(cyclenumber - 1)),
         ca199 = ifelse(ca199 == -999, NA, ca199),
         subject_id = str_pad(subject_id, 3, "left", "0"),
         lab_date = as.Date(lab_date)) %>%
  select(-c(lab_suffix, cyclenumber)) %>%
  drop_na()

# Load the clinical data
# Keep the treatment start date, imaging date and blood draw dates
clin_data <- read_excel(here("data/clinical_073123.xlsx"),
                        sheet = "CheckPAC_DATA4_OSupdate_tojh") %>%
  select(subject_id = StudySubjectID,
         tx_start_date = Nivo_StartDate,
         starts_with("IT_0_")) %>%
  rename(ct_flwup_date = IT_0_E4_C19_1)

# Fill the missing imaging date with the next available date
# this is based on my examination of the data
# Find the next available date
ct_missing_fill <- clin_data %>%
  select(subject_id, starts_with("IT_0_")) %>%
  pivot_longer(cols = -subject_id,
               names_to = "visit_id", values_to = "visit_date") %>%
  filter(visit_date != "#NULL!") %>%
  distinct(subject_id, .keep_all = TRUE)

ct_missing_fill_names <- ct_missing_fill$subject_id
ct_missing_fill <- ct_missing_fill$visit_date
names(ct_missing_fill) <- ct_missing_fill_names

# Fill the dates that are missing
clin_data <- clin_data %>%
  mutate(ct_flwup_date = ifelse(ct_flwup_date == "#NULL!",
                                ct_missing_fill[subject_id], ct_flwup_date)) %>%
  select(-starts_with("IT_0_"))

# Combine the plasma manifest with the clinical data
all_data <- inner_join(clin_data, id_map, by = "subject_id") %>%
  inner_join(mfst, by = "subject_id")

for (i in colnames(all_data)) {
  if (str_detect(i, "[Dd]ate")) {
    if (!(any(class(all_data[[i]]) == "POSIXt") || (class(all_data[[i]]) == "Date")))  {
      all_data[[i]] <- as.Date(as.numeric(all_data[[i]]), origin = "1899-12-30")
    } else {
      all_data[[i]] <- as.Date(all_data[[i]])
    }
  }
}
```


```{r days_from_tx}
# Compute days from the treatment
all_data2 <- all_data %>%
  mutate(days_from_tx = as.numeric(difftime(lab_date, tx_start_date,
                                            units = "days")),
         days_btw_tx_ct = as.numeric(difftime(ct_flwup_date, tx_start_date,
                                              units = "days")))

# Order patients based on baseline blood draw date
patient_order <- all_data2 %>%
  group_by(patient_id) %>%
  summarize(order_baseline = min(days_from_tx, na.rm = TRUE)) %>%
  ungroup() %>%
  arrange(order_baseline) %>%
  pull(patient_id)

# Select P2 dates for on-treatment
date_blp2s <- all_data2 %>%
  group_by(patient_id) %>%
  summarize(patient_id = patient_id[1],
            tx_start_date = tx_start_date[1],
            ct_flwup_date = ct_flwup_date[1],
            baseline = get_baseline(lab_date, tx_start_date),
            endpoint = get_p2(lab_date, visit_cycle),
            ct_flwup = ifelse(all(is.na(ct_flwup_date)), FALSE, TRUE)) %>%
  ungroup()

attach(date_blp2s)
dmin <- quantile(as.numeric(endpoint-tx_start_date), p=0.05, na.rm=TRUE)
dmax <- quantile(as.numeric(endpoint-tx_start_date), p=0.95, na.rm=TRUE)
detach(date_blp2s)

#dir.create("output")
#write.csv(date_blp2s, "output/checkpac_baseline_p2_dates.csv", quote=FALSE, row.names=FALSE)
```


```{r ext-fig15, fig.width = 10, fig.height = 7, dev = c("png", "pdf")}
blp2s <- date_blp2s %>%
  select(patient_id, baseline, endpoint) %>%
  pivot_longer(-patient_id,
               names_to = "timepoint", values_to = "date_blood_draw")

blp2_avail <- date_blp2s %>%
  mutate(baseline = ifelse(!is.na(baseline), TRUE, FALSE),
         endpoint = ifelse(!is.na(endpoint), TRUE, FALSE)) %>%
  mutate(keep_patient = baseline & endpoint) %>%
  #mutate(keep_patient = endpoint) %>%
  select(patient_id, keep_patient) %>%
  mutate(selection = ifelse(keep_patient, "", "X")) %>%
  mutate(patient_id = factor(patient_id, levels = patient_order),
         patient_yc = as.numeric(patient_id))

all_data3 <- all_data2 %>%
  rename(date_blood_draw = lab_date) %>%
  left_join(blp2s) %>%
  mutate(twomonth_mark = ifelse(visit_cycle == "4", 1, NA)) %>%
  mutate(patient_id = factor(patient_id, levels = patient_order),
         patient_yc = as.numeric(patient_id),
         visit_cycle = factor(visit_cycle,
                              levels = c("baseline", 1, 2, 3, 4, 5))) %>%
  mutate(selected = ifelse(timepoint %in% c("baseline", "endpoint"), 1,
                           ifelse(is.na(visit_cycle), 0, 0.3)))

ggplot() +
  geom_point(data = all_data3, aes(x = days_from_tx, y = patient_yc,
                                   color = visit_cycle)) +
  geom_point(data = all_data3, aes(x = days_from_tx, y = patient_yc,
                                   shape = twomonth_mark)) +
  scale_shape_identity() +
  geom_vline(xintercept = 0, linetype = "dashed") +
  geom_segment(data = all_data3,
               aes(x = days_btw_tx_ct, xend = days_btw_tx_ct,
                   y = patient_yc - 0.25, yend = patient_yc + 0.25),
               linetype = "dashed", color = "blue") +
  geom_rect(aes(xmin = dmin, xmax = dmax, ymin = -Inf, ymax = Inf),
            fill = "steelblue", alpha = 0.1) +
  scale_x_continuous(breaks = seq(-42, 119, 7), labels = seq(-42, 119, 7),
                     expand = c(0, 0)) +
  scale_y_continuous(breaks = sort(unique(all_data3$patient_yc)),
                     labels = levels(all_data3$patient_id),
                     sec.axis = sec_axis(~., breaks = blp2_avail$patient_yc,
                                         labels = blp2_avail$selection),
                     expand = c(0, 0)) +
  coord_cartesian(xlim = c(-42, 105)) +
  labs(x = "Days from treatment", y = "", title = NULL) +
  theme_bw() +
  theme(axis.text.y.right = element_text(color = "red"),
        panel.grid.major.x = element_line(linetype = 3),
        panel.grid.major.y = element_line(linetype = 3),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank()) +
  guides(color = guide_legend(title = "Timepoint"),
         alpha = "none")
```


