---
title: cfDNA fragmentation features reflect underlying tumor biology in pancreatic cancer  
---


Fraction of CNV changes by chromosomal arm for PDAC cancers in TCGA are shown alongside assessment of copy number changes in CheckPAC patients, as quantified by the median log-ratio by arm for tumor copy number gains and losses and z-scores by arm for all patients with plasma samples at baseline, and at follow-up for each of the clinical RECIST 1.1 response categories.


```{r packages, message=FALSE}
library(tidyverse)
library(readxl)
library(data.table)
library(RColorBrewer)
library(here)
devtools::load_all(here("code/liver.tools"))
source(here("code/functions.R"))
```


```{r data}
# Load TCGA data 
fig.data <- readRDS(here("data/fig2c_p2_data.rds"))
# Update liver to pancreas at the source

# Load CNV data from FACETS
fnames <- list.files(here("data/estimates"), pattern=".tsv$")
noares <- fnames %>%
  file.path(here("data/estimates"), .) %>%
  map_dfr(get_cnv)

# Load DELFI features
features <- read_csv(here("data/checkpac_features.csv"))
clean_data<-read_excel(here("data/supplementary_tables.xlsx"), sheet="Table S1",range="A2:X45")
delfi_pred<-read_excel(here("data/supplementary_tables.xlsx"), sheet="Table S2",range="A2:P219")
delfi_pred<-delfi_pred %>% filter(Timepoint!="follow up", Timepoint!="cycle 1")
DELFI_complete<-inner_join(delfi_pred,clean_data, by="Patient")
```


```{r fig3b, fig.height = 40, fig.width = 20,dev = c("png", "pdf")}
# TCGA panel
setDT(fig.data)
fig.data[,bin:=as.factor(rev(bin))][]
fig.data$disease<-"TCGA Pancreatic Cancers\n "
tcga <- fig.data %>%
  mutate(disease="TCGA Pancreatic Cancers\n ") %>%
  ggplot(aes(x=value, y=bin, color=change, fill=change)) +
  facet_grid(arm~disease, scales = 'free_y') +
  geom_col() +
  scale_x_continuous() +
  scale_fill_manual(values=c("#FF6636","#66CCCC"))+
  scale_colour_manual(values=c("#FF6636","#66CCCC"))+
  theme_minimal(base_size=24) +
  theme(strip.text.y = element_text(angle = 0,
                                    hjust = 0.5,
                                    vjust = 0.5,
                                    size=20),
        strip.text.x=element_text(size=20),
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        panel.grid =  element_blank(),
        legend.title = element_blank(),
        legend.position="bottom") +
  xlab("Proportion of\ncases with CNV") +
  ylab("Chromosome Arm")

# Tumor panel 
noares1 <- noares %>%
    mutate(colors= case_when((cnv_score>=.05) ~ "Positive",
    (cnv_score>=(-.05)) ~ "Neutral",
    (cnv_score<(-.05)) ~ "Negative"))
noares1$colors<-factor(noares1$colors, levels=c("Positive","Neutral","Negative"))
noares1 <- noares1 %>% mutate(transp=if_else(colors=="Neutral",1,1))
noares1$y=1
noares1<-noares1 %>% rename(Arm= chrarm)
noares1  <- noares1 %>%rename(id=ID)
noares1<-noares1 %>% mutate(BOR="Tumor")
noares1  <- noares1 %>%
        rename(z=cnv_score) %>%
        mutate(root2=nroot(z, 2)) %>%
        mutate(root3=nroot(z, 3))
noares2<-noares1 %>% mutate(source="Tumor")

# Plasma panel 
DELFI_complete$id<-DELFI_complete$Sample
DELFI_complete<-DELFI_complete %>% mutate(BOR = case_when((Timepoint == "Baseline") ~ "Baseline",
                                                          TRUE ~ `BOR RECIST 1.1`))
DELFI_complete<-DELFI_complete %>% filter(`BOR RECIST 1.1`!="Not Evaluable")

cols <- c("#FF6636","#66CCCC","#FFCC66","#FFCC66","#66CCCC","#FFFF00","#FF9900")
features <- features %>% select(id,starts_with("z"))
features <- inner_join(features, DELFI_complete %>%
                       select(id,BOR),
                   by=c("id"="id"))
features <- features %>%
  gather(key="Arm", value="Zscore", starts_with("z"))
features$Arm <- sapply( str_split(features$Arm,"_"), "[", 2 )
chrlevels <- c("1p","1q","2p","2q","3p","3q","4p","4q","5p",
               "5q","6p","6q","7p","7q","8p","8q","9p","9q",
               "10p","10q","11p","11q","12p","12q",
               "13q","14q","15q","16p","16q","17p","17q",
               "18p","18q","19p","19q","20p","20q","21q","22q")
features$Arm <- factor(features$Arm,
                       levels = chrlevels)
features <- features %>% mutate(colors=if_else(Zscore<=0,"Negative","Positive"))
features <- features %>%
        mutate(colors=if_else(Zscore>= -3.22 & Zscore <= 3.22 ,
                              "Neutral", features$colors))
features <- features %>% mutate(transp=if_else(colors=="Neutral",1,1))

features$colors<-factor(features$colors,levels=c("Positive","Negative","Neutral"))
features$y=1

features <- features %>% mutate(Zscore=if_else(Zscore<=-150,-150,Zscore))
features <- features %>% mutate(Zscore=if_else(Zscore>150,150,Zscore))
features2  <- features %>%
        rename(z=Zscore) %>%
        mutate(root2=nroot(z, 2)) %>%
        mutate(root3=nroot(z, 3))
features2<-features2 %>%mutate(source="Plasma")

# Combine tumor and plasma data
features3<-rbind(features2, noares2) %>%
  mutate(BOR = factor(BOR, levels = c("Tumor", "Baseline", "Progressive Disease",
                                      "Stable Disease", "Partial Response")))
# Restrict data to -150 to 150 for plasma and -1.5 to 1.5 for tumor
xaxis_lim <- tibble(id = paste(rep("DUMMY", 10), 1:10),
                    BOR = factor(rep(levels(features3$BOR), each = 2),
                                 levels = levels(features3$BOR)),
                    Arm = factor("1p", levels = chrlevels),
                    colors = factor(rep(c("Negative", "Positive"), 5),
                                    levels = levels(features3$colors)),
                    z = c(c(-1.5, 1.5),
                          rep(c(-150, 150), 4)),
                    transp = 0,
                    y = 1)
ns <- features3 %>%
    group_by(BOR) %>%
    summarize(n=length(unique(id))) %>%
    mutate(label=paste0(BOR, "\n(n = ", n, ")"))
ns <- setNames(ns$label, ns$BOR)

jhu <- features3 %>%
    bind_rows(xaxis_lim) %>%
    ggplot(aes(x=y, y=z, color=colors, alpha=transp)) +
    geom_point(size=5) +
    facet_grid(Arm ~ BOR, labeller=labeller(BOR=ns), scales = "free_x") +
    coord_flip() +
    theme_minimal(base_size=24) +
    theme(strip.text.y.left = element_text(angle = 0,
                                           hjust = 0.5,
                                           vjust = 0.5),
          axis.text.y = element_blank(),
          axis.ticks = element_blank(),

          panel.grid =  element_blank(),
          plot.title = element_blank(),
          legend.title = element_blank(),
          axis.title.y=element_blank(),
          legend.position = "bottom",
          strip.text.y = element_blank()) +
    scale_y_continuous() +
    scale_alpha_identity() +
    scale_fill_manual(values = cols) +
    scale_color_manual(values = cols) +
    ylab("Z-Score")

cowplot::plot_grid(tcga, jhu, nrow = 1, rel_widths = c(1, 2.5), align = "h")
```
