---
title: Multivariate hazard analyses demonstrate on-treatment ARTEMIS-DELFI scores as independent predictors of overall survival for patients in the CheckPAC trial
---


Multivariate Cox proportional hazard analyses were generated for each molecular method and fit to overall survival adjusting for clinical subgroups. Each of the indicated subgroups that have been shown to be significant on univariate analyses in previous studies (7) have been included in the multivariate analysis.  Hazard models are shown using ARTEMIS-DELFI scores at baseline and after one cycle of treatment for patients in the CheckPAC study. 


```{r packages, message = FALSE}
library(here)
library(readxl)
library(survival)
source(here("code/hr_plot_funcs.R"))
```


```{r load_clin_data}
baseline_c <- read_excel("data/CheckPAC_clinical_annotated011622.xlsx") %>%
  select(subject_id = StudySubjectID,
         age,
         weightloss5_yn = weighttloss5_yn,
         ecog_ps = PS_cycle1,
         bor = `Best overall confirmed response (RECIST)`,
         ca199 = LAB_CA199_E1_C4,
         crp = LAB_CRP_E1_C4,
         bilirubin = LAB_Bilirubin_E1_C4,
         albumin = LAB_Albumin_E1_C4) %>%
  mutate(mgps = case_when(crp <= 10 ~ 0,
                          crp > 10 & albumin >= 35 ~ 1,
                          crp > 10 & albumin < 35 ~ 2))
long_c <- read_excel("data/clinical_073123.xlsx") %>%
  select(subject_id = StudySubjectID,
         arm = Arm,
         nivo_start_date = Nivo_StartDate,
         pfs_date = PFS_date,
         surv_date = Survival_date,
         surv_status = Survival_status) %>%
  mutate(surv_status = case_when(surv_status == 2 ~ "Dead",
                                 surv_status == 1 ~ "Alive"),
         arm = factor(arm))
clin_c <- inner_join(baseline_c, long_c, by = "subject_id")

# Fix the date formats
for (i in colnames(clin_c)) {
  if (str_detect(i, "[Dd]ate")) {
    if (!any(class(clin_c[[i]]) == "POSIXt")) {
      clin_c[[i]] <- as.Date(as.numeric(clin_c[[i]]), origin = "1899-12-30")
    } else {
      clin_c[[i]] <- as.Date(clin_c[[i]])
    }
  }
}

# Compute overall and progression-free survival time
surv_c <- clin_c %>%
  mutate(pfs = as.numeric(difftime(pfs_date, nivo_start_date)),
         os = as.numeric(difftime(surv_date, nivo_start_date))) %>%
  mutate(death = ifelse(surv_status == "Dead", 1, 0))
```

```{r load_artemis_delfi}
artemis_delfi_c <- read_csv("data/Test_set_pacto.csv") %>%
  select(id, score, model) %>%
  filter(model == "ARTEMIS_single_DELFI_SSLs_Ensemble")

# Select baseline and two-week visit timepoints
blood_draw_c <- read_excel("data/checkpac_plasma_manifest.xlsx") %>%
  select(cgid = `Sample ID*`, patient_id = Patient, date_blood_draw = Date) %>%
  mutate(date_blood_draw = as.Date(date_blood_draw))
# See "/dcs05/scharpf/data/skoul/Projects/checkpac/patient_selection" for more
# information
blood_draw_c[blood_draw_c$cgid == "CGPLPA223P16",
             "date_blood_draw"] <- as.Date("2019-12-20")
blood_draw_c[blood_draw_c$cgid == "CGPLPA794P4",
             "date_blood_draw"] <- as.Date("2020-02-21")
blood_draw_bl_or_p1_c <- blood_draw_c %>%
  mutate(visit = str_extract(cgid, "(?<=CG[A-Z]{1,5}[0-9]{1,5}P).*"),
         visit = sapply(visit, function(x) str_split(x, "_")[[1]][1])) %>%
  filter(visit == "" | visit == "1")
blood_draw_bl_and_p1_c <- blood_draw_bl_or_p1_c %>%
  filter(patient_id %in% pull(filter(count(group_by(blood_draw_bl_or_p1_c,
                                                    patient_id)),
                                     n > 1),
                              patient_id))

# Join the scores with timepoint
artemis_delfi_blp1_c <- inner_join(artemis_delfi_c, blood_draw_bl_and_p1_c,
                                   by = join_by(id == cgid)) %>%
  mutate(visit = ifelse(visit == "", "baseline", "week2"))
```


```{r combine_clin_artemis_delfi}
exclude_c <- c("CGPLPA225", "CGPLPA783", "CGPLPA795")
id_map_c <- read_excel("data/ID_comparison.xlsx") %>%
  filter(!unlist(lapply(CGID, function(x) any(str_detect(x, exclude_c))))) %>%
  select(patient_id = CGID, subject_id = StudySubjectID) %>%
  mutate(patient_id = substring(patient_id, 1, 9))

surv_artemis_delfi_c <- surv_c %>%
  inner_join(id_map_c) %>%
  inner_join(artemis_delfi_blp1_c) %>%
  select(subject_id, patient_id, score, model,
         arm, ca199, bilirubin, age, mgps, ecog_ps, weightloss5_yn, visit,
         pfs, os, death) %>%
  pivot_wider(names_from = visit, values_from = score) %>%
  mutate(bilirubin_bin = bilirubin > 25,
         age_bin = age <= 65,
         mgps_bin = factor(mgps, levels = c(0, 1, 2)),
         ecog_ps_bin = ecog_ps == 1,
         ca199_bin = ca199 > 37,
         weightloss5_yn_bin = weightloss5_yn == 1)
```


```{r fig7b, fig.width = 12, fig.height = 5, dev = c("png", "pdf")}
var_labels <- c("Age, years",
                "> 65",
                "\u2264 65",
                "ECOG PS",
                "0",
                "1",
                "Weight loss, %",
                "< 5",
                "\u2265 5",
                "MGPS",
                "0",
                "1",
                "2",
                "Treatment",
                "RT+Nivo",
                "RT+Nivo/Ipi",
                "ARTEMIS DELFI score",
                "Baseline",
                "ARTEMIS DELFI score",
                "Cycle 1")

os_mf_c <- coxph(Surv(os, death) ~ age_bin + ecog_ps_bin +
                   weightloss5_yn_bin +
                   mgps_bin + arm + baseline + week2,
                 surv_artemis_delfi_c)
os_cox_tbl <- get_hr_tbl(surv_artemis_delfi_c, os_mf_c)
print(plot_hr(os_cox_tbl, var_labels = var_labels))
```


```{r fig7b-pfs, fig.width = 12, fig.height = 5, dev = c("png", "pdf"), eval = FALSE}
pfs_mf_c <- coxph(Surv(pfs, death) ~ age_bin + ecog_ps_bin +
                    weightloss5_yn_bin +
                    mgps_bin + arm + baseline + week2,
                  surv_artemis_delfi_c)

pfs_cox_tbl <- get_hr_tbl(surv_artemis_delfi_c, pfs_mf_c)
print(plot_hr(pfs_cox_tbl, var_labels = var_labels))
```
