---
title: Example of a molecular responder and non-responder to treatment with different methodologies 
---


Response to treatment is shown for a patient with clinical BOR of partial response (A) and stable disease (B). Patient clinical paths are shown in the top panels, followed by methodologies monitor response to treatment. For the top panels, patient treatment is indicated in the top two rows, radiologic assessments are plotted in the third row and last follow up is indicated in the fourth row.  Sum of target lesions were assessed from standard of care CT scans. WGMAF, targeted MAF, and ARTEMIS-DELFI score are plotted for each of the blood draw timepoints. CA19-9 was measured clinically at regular intervals throughout treatment. 


```{r packages, message = FALSE}
library(here)
library(tidyverse)
library(readxl)
library(patchwork)
```


```{r load_data}
# Reading in the ID conversion table
id.df<-read_excel(here("data/i07_input.xlsx"), sheet="ID_comparison", na = "#NULL!")
id.df$patient_id <- gsub("P$|P1$", "", id.df$CGID)
id.df$StudySubjectID<-as.numeric(id.df$StudySubjectID)

# Reading in a clinical metadata file and extracting the date of radiation, ipi, and nivo
clin.df<-read_excel(here("data/i07_input.xlsx"), sheet="CheckPAC_DATA4", na = "#NULL!")

clin.df$StudySubjectID<-as.numeric(clin.df$StudySubjectID)
clin.df$Survival_status<-as.numeric(clin.df$Survival_status)

tx_dates.df <- left_join(id.df, clin.df, by = "StudySubjectID") %>%
                 mutate(immunotx_start_date1 = pmin(Nivo_StartDate, Ipi_StartDate, na.rm = T),
                        immunotx_end_date1 = pmax(Nivo_LastDate, Ipi_LastDate, na.rm = T),
                        immunotx_start_date2 = pmin(Nivo_ReStart, Ipi_ReStart, na.rm = T),
                        immunotx_end_date2 = pmax(Nivo_ReStop, Ipi_ReStop, na.rm = T)) %>%
                 select(patient_id, radiation_date = Radiation_Date,
                        immunotx_start_date1, immunotx_end_date1, immunotx_start_date2, immunotx_end_date2) %>%
                 mutate(across(contains("date"), as.Date))
# Extracting date of death
survival.df <- left_join(id.df, clin.df, by = "StudySubjectID") %>%
                 select(patient_id, date = Survival_date, status = Survival_status) %>%
                 mutate(date = as.Date(date),
                        status = ifelse(status == 2, "Deceased", "Alive")) %>%
                 mutate(status = factor(status, levels = c("Alive", "Deceased"))) %>%
              mutate(date = as.Date(date, format = "%m/%d/%Y"))

# Reading in TMB for 34 samples with T/N/plasma WGS sequencing
tmb.df <- read_excel(here("data/i07_input.xlsx"), sheet="tumor_tmb", na = "#NULL!") %>%
            mutate(summary = paste0("Number of SBS mutations by WGS in tumor tissue (unfiltered, filtered): ", sbs.tmb, ", ", sbs.tmb.filtered))

# Reading in the patient samples summary that indicates if T/N/plasma WGS is available for each patient
wgs_patient_sample_summary.df <- read_excel(here("data/i07_input.xlsx"), sheet="patient_summary", na = "#NULL!") %>%
                                   mutate(summary = paste0("WGS data: Tumor (", tumor_tissue, "), Normal (", normal_tissue, "), Plasma (", plasma_100bp, " timepoints)"))

# Reading in DELFI scores
delfi.df <- read_excel(here("data/i07_input.xlsx"), sheet="checkpac_preds", na = "#NULL!") %>%
              rename(plasma_id = id, patient_id = Patient, score = 'ARTEMIS-DELFI Score') %>%
                mutate(date = as.Date(date, format = "%m/%d/%Y"))

# Reading in WGS MAFs
wgs_maf.df <- read_excel(here("data/i07_input.xlsx"), sheet="plasma_mafs", na = "#NULL!") %>%
                rename(date = "date_blood_draw") %>% mutate(date = as.Date(date, format = "%m/%d/%Y")) %>%
  filter(plasma_read_length!="150")

# Reading in targeted sequencing MAFs
targeted_maf.df <- read_excel(here("data/i07_input.xlsx"), sheet="targeted_sequencing", na = "#NULL!") %>%
                     mutate(date = as.Date(date_blood_draw, format = "%m/%d/%Y")) %>%
                     select(patient_id, date, mutation, maf)

# Reading in CA-19-9 measurements
ca19.df <- read_excel(here("data/i07_input.xlsx"), sheet="CA19-9", na = "#NULL!") %>%
             left_join(., id.df, by = c("biopac_id" = "Biopac ID")) %>%
             mutate(date = as.Date(date_blood_draw, format = "%m/%d/%Y")) %>%
             select(patient_id, date, ca19_9)

# Reading in RECIST response information
recist.df <- read_excel(here("data/i07_input.xlsx"), sheet="RECIST", na = "#NULL!") %>%
               left_join(., id.df, by = c("biopac_id" = "Biopac ID")) %>%
               mutate(date = as.Date(date_ct, format = "%m/%d/%Y")) %>%
               select(patient_id, date, sum_target_lesions, recist_response) %>%
               mutate(recist_response = ifelse(is.na(recist_response), "Baseline", recist_response)) %>%
               mutate(recist_response = factor(recist_response, levels = c("Baseline", "PR", "SD", "PD")))
recist.df$sum_target_lesions<-as.numeric(recist.df$sum_target_lesions)

# Getting the dates of the first / last measurements to be plotted for each patient
date_range.df <- bind_rows(pivot_longer(tx_dates.df, !patient_id, names_to = NULL, values_to = "date", values_drop_na = T),
                           select(survival.df, c(patient_id, date)),
                           select(delfi.df, c(patient_id, date)),
                           select(wgs_maf.df, c(patient_id, date)),
                           select(targeted_maf.df, c(patient_id, date)),
                           select(ca19.df, c(patient_id, date)),
                           select(recist.df, c(patient_id, date))) %>%
                   group_by(patient_id) %>%
                   summarise(date_initial = min(date),
                             date_final = max(date))
```


```{r fig8, fig.width = 11, fig.height = 14, dev = c("png", "pdf")}
# Generating a figure of cfDNA  metrics, ca19-9,  and clinical information for each patient
exampleA <- id.df$patient_id[str_detect(id.df$patient_id, "223")]
exampleB <- id.df$patient_id[str_detect(id.df$patient_id, "220")]
patient_ids <- c(exampleA, exampleB)

for (i in 1:length(patient_ids)) {

  date_range.df.i <- filter(date_range.df, patient_id == patient_ids[i])
  date_initial.i <- date_range.df.i$date_initial
  date_final.i <- date_range.df.i$date_final

  # Clinical information
  recist.df.i <- filter(recist.df, patient_id == patient_ids[i])
  recist.df.i$recist_response<-as.factor(recist.df.i$recist_response)
  tx_dates.df.i <- filter(tx_dates.df, patient_id == patient_ids[i])
  survival.df.i <- filter(survival.df, patient_id == patient_ids[i])
  survival.df.i$status<-as.factor(survival.df.i$status)
  
  p.clinical <- ggplot() +
                  geom_point(data = NULL, aes(x = tx_dates.df.i$radiation_date, y = 1.75), color = "black", shape = 4, size = 4) +
                  geom_rect(data = NULL, aes(xmin = tx_dates.df.i$immunotx_start_date1, xmax = tx_dates.df.i$immunotx_end_date1, ymin = 1.3125, ymax = 1.4375), color = NA, fill = "steelblue", alpha = 0.5) +
                  geom_rect(data = NULL, aes(xmin = tx_dates.df.i$immunotx_start_date2, xmax = tx_dates.df.i$immunotx_end_date2, ymin = 1.3125, ymax = 1.4375), color = NA, fill = "steelblue", alpha = 0.5) +
                  geom_point(data = survival.df.i, aes(x = date, y = 0.625, shape = status), size = 4) +
                  cowplot::theme_cowplot() +
                  theme(axis.title.y = element_blank(),
                        axis.line.y = element_blank(),
                        axis.ticks.y = element_blank(),
                        axis.title.x = element_blank()) +
                  scale_color_manual(name = "RECIST 1.1", 
                                     breaks = c("Baseline", "PR", "SD", "PD"),
                                     values  = c("grey65", "#449c52", "#e8b82a", "#9c3b3b"),
                                     drop = F) +
                  scale_shape_manual(name = "Survival at last follow up",
                                     breaks = c("Alive", "Deceased"),
                                     values = c(2, 6), 
                                     drop = F) +
                  xlab("Date") +
                  coord_cartesian(xlim = c(date_initial.i, date_final.i), ylim = c(0.375, 2)) +
                  scale_y_continuous(breaks = c(0.625, 1, 1.375, 1.75), labels = c("Last follow up", "CT scan", "Immunotherapy", "Radiation")) +
                  scale_x_date(position = "top")

  # Add CT scan / RECIST information for patients that have it
  if (nrow(recist.df.i) > 0) {
    p.clinical <- p.clinical +
                    geom_point(data = recist.df.i, 
                               aes(x = date, y = 1, color = recist_response),
                               size = 4
                           #    ,show.legend = FALSE
                               ) + #ADDED-LG
                    guides(color = guide_legend(order = 1), shape = guide_legend(order = 2))
  }

  # Sum target lesions
  p.sum_target_lesions <- ggplot(recist.df.i, aes(x = date, y = sum_target_lesions)) +
                            geom_line() +
                            geom_point() +
                            cowplot::theme_cowplot() +
                            theme(axis.title.y = element_text(angle = 0, vjust = 0.5, margin = margin(r = -45, unit = "pt")),
                                  axis.title.x = element_blank(),
                                  axis.ticks.x = element_blank(),
                                  axis.text.x = element_blank()) +
                           ylab("Sum of\ntarget\nlesions\n(mm)") +
                           xlab("Date") +
                           coord_cartesian(xlim = c(date_initial.i, date_final.i))

  # ARTEMIS-DELFI scores
  delfi.df.i <- filter(delfi.df, patient_id == patient_ids[i])
  p.delfi <- ggplot(delfi.df.i, aes(x = date, y = score)) +
               geom_line() +
               geom_point() +
               cowplot::theme_cowplot() +
               theme(axis.title.y = element_text(angle = 0, vjust = 0.5, margin = margin(r = -45, unit = "pt")),
                     axis.title.x = element_blank(),
                     axis.ticks.x = element_blank(),
                     axis.text.x = element_blank()) +
               ylab("ARTEMIS-DELFI score") +
               xlab("Date") +
               coord_cartesian(xlim = c(date_initial.i, date_final.i), ylim = c(0, 1)) +
               scale_y_continuous(breaks = c(0, 0.2, 0.4, 0.6, 0.8, 1), labels = c("0", "0.2", "0.4", "0.6", "0.8", "1"))

  # WGS MAFs
  wgs_maf.df.i <- filter(wgs_maf.df, patient_id == patient_ids[i])
  p.wgs_maf <- ggplot(wgs_maf.df.i, aes(x = date, y = mf * 100)) +
                 geom_line() +
                 geom_point() +
                 cowplot::theme_cowplot() +
                 theme(axis.title.y = element_text(angle = 0, vjust = 0.5, margin = margin(r = -45, unit = "pt")),
                       axis.title.x = element_blank(),
                       axis.ticks.x = element_blank(),
                       axis.text.x = element_blank()) +
                 ylab("WGMAF\n(%)") +
                 xlab("Date") +
                 coord_cartesian(xlim = c(date_initial.i, date_final.i))

  # Targeted MAFs
  targeted_maf.df.i <- filter(targeted_maf.df, patient_id == patient_ids[i])
  p.targeted_maf <- ggplot(targeted_maf.df.i, aes(x = date, y = maf * 100, group = mutation, color = mutation)) +
                      geom_line() +
                      geom_point() +
                      cowplot::theme_cowplot() +
                      theme(axis.title.y = element_text(angle = 0, vjust = 0.5, margin = margin(r = -45, unit = "pt")),
                            axis.title.x = element_blank(),
                            axis.ticks.x = element_blank(),
                            axis.text.x = element_blank()) +
                      scale_color_brewer(name = "Mutation", palette = "Dark2") +
                      ylab("Targeted MAF\n(%)") +
                      xlab("Date") +
                      coord_cartesian(xlim = c(date_initial.i, date_final.i))

  # CA-19-9 levels
  ca19.df.i <- filter(ca19.df, patient_id == patient_ids[i])
  p.ca19 <- ggplot(ca19.df.i, aes(x = date, y = ca19_9)) +
              geom_point() +
              geom_line() +
              cowplot::theme_cowplot() +
              theme(axis.title.y = element_text(angle = 0, vjust = 0.5, margin = margin(r = -45, unit = "pt")),
                    axis.title.x = element_blank())+
              ylab("CA 19-9\n(U/mL)") +
              xlab("Date") +
              coord_cartesian(xlim = c(date_initial.i, date_final.i))

   # Plot title information
   tmb.df.i <- filter(tmb.df, patient_id == patient_ids[i])
   wgs_patient_sample_summary.df.i <- filter(wgs_patient_sample_summary.df, patient_id == patient_ids[i])
   plot.title.i <- paste0(patient_ids[i], "\n",
                          ifelse(nrow(wgs_patient_sample_summary.df.i) > 0, wgs_patient_sample_summary.df.i$summary, "WGS data: Tumor (No), Normal (No), Plasma (0 timepoints)"), "\n",
                          ifelse(nrow(tmb.df.i) > 0, tmb.df.i$summary, "Number of SBS mutations by WGS in tumor tissue (unfiltered, filtered): not measured"))

   p.patient.i <- p.clinical / p.sum_target_lesions /p.wgs_maf / p.targeted_maf / p.delfi /  p.ca19 +
                   plot_layout(heights = c(1.15, 0.75, 0.75, 0.75, 0.75, 0.75)) +
                   plot_annotation(title = plot.title.i,
                                   theme = theme(plot.title = element_text(hjust = 0.5, size = 14)))

  print(p.patient.i)
}
```
