---
title: Multivariate hazard analyses demonstrate on-treatment ARTEMIS-DELFI scores as independent predictors of overall survival for patients in the PACTO trial
---


Multivariate Cox proportional hazard analyses were generated for each molecular method and fit to overall survival adjusting for clinical subgroups. Each of the indicated subgroups that have been shown to be significant on univariate analyses have been included in the multivariate analysis.  Hazard models are shown using ARTEMIS-DELFI scores at baseline and 8-week follow-up for patients in the PACTO study 


```{r packages, message = FALSE}
library(here)
library(readxl)
library(survival)
source(here("code/hr_plot_funcs.R"))
```


```{r load_clin_data}
clin_p <- read_excel(here("data/PACTO_BL_EOT_EOS_PFS_TA_jh.xlsx")) %>%
  select(subject_id = StudySubjectID,
         age,
         #weightloss5_yn is not available
         arm = INCL_120_E1_C1,
         #ecog_ps is not available
         tx_start_date = NABDATE_E2_1_C2,
         bor = BESTresponse,
         pfs_date = PFS_date,
         surv_date = DEATHDATE_E4_C9,
         surv_status = SURVIVAL_E4_C9,
         ca199 = CA19_9,
         crp = CRP,
         bilirubin = bili,
         hemoglobin,
         albumin = Albumin,
         mgps = mGPS) %>%
  mutate(arm = ifelse(arm == 9, 1, arm),
         arm = factor(arm)) # 9 was safety cohort for arm-1
# arm-1: CT+Nab+Tocili
# arm-2: CT+Nab

# Fix the date formats
for (i in colnames(clin_p)) {
  if (str_detect(i, "[Dd]ate")) {
    if (!any(class(clin_p[[i]]) == "POSIXt")) {
      clin_p[[i]] <- as.Date(as.numeric(clin_p[[i]]), origin = "1899-12-30")
    } else {
      clin_p[[i]] <- as.Date(clin_p[[i]])
    }
  }
}

# Compute overall and progression-free survival time
surv_p <- clin_p %>%
  mutate(pfs = as.numeric(difftime(pfs_date, tx_start_date)),
         os = as.numeric(difftime(surv_date, tx_start_date))) %>%
  mutate(death = ifelse(surv_status == 2, 1, 0))
```


```{r load_artemis_delfi}
artemis_delfi_p <- read_csv(here("data/Test_set_pacto.csv")) %>%
  select(id, artemis_delfi_score = score, model) %>%
  filter(model == "ARTEMIS_single_DELFI_SSLs_Ensemble")

# Select baseline and two-week visit timepoints
blood_draw_p <- read_excel(here("data/PACTO Masterfile.xlsx")) %>%
  select(cgid = `CGID...3`, date_blood_draw = `Date of draw`,
         subject_id = PACTO) %>%
  mutate(date_blood_draw = as.Date(date_blood_draw),
         patient_id = str_extract(cgid, "^CG[A-Z]{1,5}[0-9]{1,5}"),
         subject_id = str_pad(subject_id, 3, "left", "0"))
blood_draw_bl_or_p1_p <- blood_draw_p %>%
  filter(!str_detect(cgid, "89[4,5,6]P.*_.*")) %>% # removing duplicate samples
  filter(!str_detect(cgid, "868P.*_.*")) %>% # removing duplicate samples
  filter(!str_detect(cgid, "892")) %>% # removing sample with missing score
  mutate(visit = str_extract(cgid, "(?<=CG[A-Z]{1,5}[0-9]{1,5}P).*"),
         visit = sapply(visit, function(x) str_split(x, "_")[[1]][1])) %>%
  filter(visit == "" | visit == "2")
blood_draw_bl_and_p1_p <- blood_draw_bl_or_p1_p %>%
  filter(patient_id %in% pull(filter(count(group_by(blood_draw_bl_or_p1_p,
                                                    patient_id)),
                                     n > 1),
                              patient_id))

# Join the scores with timepoint
artemis_delfi_blp1_p <- inner_join(artemis_delfi_p, blood_draw_bl_and_p1_p,
                                   by = join_by(id == cgid)) %>%
  mutate(visit = ifelse(visit == "", "baseline", "week8"))
```


```{r combine_clin_artemis_delfi}
surv_artemis_delfi_p <- surv_p %>%
  inner_join(artemis_delfi_blp1_p) %>%
  select(subject_id, patient_id, artemis_delfi_score, model,
         arm, ca199, bilirubin, age, mgps, #ecog_ps, weightloss5_yn,
         visit,
         pfs, os, death) %>%
  pivot_wider(names_from = visit, values_from = artemis_delfi_score) %>%
  mutate(bilirubin_bin = bilirubin > 25,
         age_bin = age <= 65,
         mgps_bin = factor(mgps, levels = c(0, 1, 2)),
         ca199_bin = ca199 > 37) %>%
  drop_na()
```


```{r fig9a, fig.width = 12, fig.height = 5, dev = c("png", "pdf")}
var_labels <- c("Age, years",
                "> 65",
                "\u2264 65",
                #"ECOG\nPerformance Status",
                #"0",
                #"1",
                #"Weight loss, %",
                #"< 5",
                #"\u2265 5",
                "MGPS",
                "0",
                "1",
                "2",
                "Bilirubin, \u00b5mol/L",
                "\u2264 25",
                "> 25",
                "Treatment",
                "CT+Nab+Tocili",
                "CT+Nab",
                "ARTEMIS DELFI score",
                "Baseline",
                "ARTEMIS DELFI score",
                "On-treatment (week 8)")

os_mf_p <- coxph(Surv(os, death) ~ age_bin + #ecog_ps_bin +
                   #weightloss5_yn_bin +
                   mgps_bin + bilirubin_bin + arm + baseline + week8,
                 surv_artemis_delfi_p)
os_pox_tbl <- get_hr_tbl(surv_artemis_delfi_p, os_mf_p)
print(plot_hr(os_pox_tbl, var_labels = var_labels))
```


```{r fig9a-pfs, fig.width = 12, fig.height = 5, dev = c("png", "pdf"), eval = FALSE}
pfs_mf_p <- coxph(Surv(pfs, death) ~ age_bin + #ecog_ps_bin +
                    #weightloss5_yn_bin +
                    mgps_bin + bilirubin_bin + arm + baseline + week8,
                  surv_artemis_delfi_p)

pfs_pox_tbl <- get_hr_tbl(surv_artemis_delfi_p, pfs_mf_p)
print(plot_hr(pfs_pox_tbl, var_labels = var_labels))
```
